Dynamics of thin liquid film flows studied with the lattice Boltzmann method

### 2023-07-24 chapters/Paper_two.tex
Chapter two feedback Jens

Coating processes heavily rely on the affinity between a fluid and a substrate~\cite{RevModPhys.81.739}.

@@

leading to film rupture~\cite{oronLongscaleEvolutionThin1997, RevModPhys.81.1131} or wrinkling~\cite{dasilvasobrinhoStudyDefectsUltrathin1999}. 

@@

The advantage of such approach in comparison with molecule-resolved methods such as Molecular Dynamics~\cite{haile1992molecular, PhysRevE.100.023108, doi:10.1063/1.1290698, grabow1988thin}

@@

For instance, in the context of dewetting, rupture times measured in experiments are shorter than predicted by simulations of the thin-film equation~\cite{herminghaus1996dewetting,herminghaus1998dewetting,becker2003complex}.
More in general, thermal fluctuations can accelerate the appearance of film instabilities~\cite{Rauscher2008,tsekov1993effect, PhysRevLett.99.114503, PhysRevE.100.023108}.
The first theoretical and experimental studies, by means of light scattering techniques, on film rupture influenced by thermal fluctuations date back to the 60s~\cite{doi:10.1021/ja01014a015}.
Almost half a century later, a stochastic version of Eq.~(\ref{eq:thin_film}) was derived~\cite{Grun2006, Mecke_2005, PhysRevLett.95.244505}, through a lubrication approximation of the Navier-Stokes-Landau-Lifshitz 
equations of fluctuating hydrodynamics~\cite{Landau1987Fluid}.
A strong agreement between experiments and the stochastic thin-film equation was found in showing that the variance of the interfacial roughness cannot be explained without adding a fluctuating term~\cite{PhysRevLett.99.114503}.
The influence of thermal fluctuations on the short and long time morphology of a film has been studied with simulations of the one and two dimensional stochastic thin-film equations~\cite{PhysRevE.92.061002, alizadeh_pahlavan_cueto-felgueroso_hosoi_mckinley_juanes_2018}. 
Recently, new theoretical insights have been gained explaining film rupture as a consequence of the combined action of thermal fluctuations and drainage, both numerically~\cite{shah_van_steijn_kleijn_kreutzer_2019} and experimentally~\cite{PhysRevLett.125.158001}.

@@

\section{Linear Stability Analysis of the Stochastic Thin-Film Equation}\label{sec:stoch_thin_film}
The one-dimensional stochastic thin-film equation reads~\cite{Grun2006, Mecke_2005}:
\begin{equation}\label{eq:stoch_thin_film_simple}
    \partial_t h = \partial_x \left[\frac{h^3}{3\mu}\partial_x p + \sqrt{\frac{2k_BTh^3}{3\mu}}\mathcal{N}\right],
\end{equation}
where $k_B$ is Boltzmann's constant, $T$ is the temperature and $\mathcal{N}$ is a Gaussian white noise with
\begin{align}\label{eq:correlation}
    \langle\mathcal{N}(x,t)\rangle =&~0,\\ 
    \quad \langle\mathcal{N}(x,t)\mathcal{N}(x',t')\rangle =&~\delta(x-x')\delta(t-t').
\end{align}
In order to understand the influence of thermal fluctuations on the temporal evolution of the film height, we cursory recall the linear stability analysis of Eq.~(\ref{eq:stoch_thin_film_simple}) 
\cite{PhysRevE.100.023108, PhysRevE.93.013120}.
Introducing the deviation $\delta h$ from the mean height $h_0$, such that $h = h_0 + \delta h$ and $\delta h \ll h_0$, 
the linearised stochastic thin-film equation is obtained
\begin{equation}\label{eq:linearstability_realspace}
    \partial_t \delta h = \frac{h_0^3}{3\mu}\Big(\Pi^{\prime}(h_0)\partial_x^2\delta h - \gamma \partial_x^4\delta h\Big) + \sqrt{\frac{2k_BT h_0^3}{3\mu}} \partial_x \mathcal{N},
\end{equation}
where $\Pi^{\prime}(h_0) \equiv \frac{\partial \Pi(h)}{\partial h}\bigg\rvert_{h=h_0}$.
Here, it is assumed that also the noise is small, in the sense that $k_B T \ll \gamma h_0^2$.
To derive a dispersion relation for this system, first the Fourier transforms of the perturbation $\delta h$  
\begin{equation}\label{eq:fourier_delta_h}
    \tilde{\delta h}(q,t) = \int_{-\infty}^{\infty} \delta h(x,t) e^{-ixq} dx,
\end{equation}
and the noise term $\mathcal{N}$
\begin{equation}\label{eq:fourier_noise}
    %\tilde{\mathcal{N}}(\mathbf{q},t) = \int_{-\infty}^{\infty} \mathcal{N}(\mathbf{x},t) e^{-i\mathbf{x}\cdot\mathbf{q}} \dd[2]{\mathbf{x}}.
    \tilde{\mathcal{N}}(q,t) = \int_{-\infty}^{\infty} \mathcal{N}(x,t) e^{-ixq} dx.
\end{equation}
must be performed.

Inserting $\tilde{\delta h}$ and $\tilde{\mathcal{N}}$ in~(\ref{eq:linearstability_realspace}) yields
\begin{equation}\label{eq:stability_fourier}
    \partial_t \tilde{\delta h} = \omega(q)\tilde{\delta h} +i\sqrt{\frac{2k_BTh_0^3}{3\mu}}q\tilde{\mathcal{N}},
\end{equation}
with dispersion relation $\omega(q)$ given by
\begin{equation}\label{eq:dispersion}
    \omega(q) = \frac{1}{t_0}\left[2\left(\frac{q}{q_0}\right)^2 - \left(\frac{q}{q_0}\right)^4\right].
\end{equation}
The maximum, $q_0$, of $\omega(q)$ and the characteristic time, $t_0$, are specific to the substrate the film is deposited on and their expressions are~\cite{PhysRevLett.99.114503}
\begin{equation}\label{eq:q0}
    q_0^2 = \frac{1}{2\gamma}\frac{\partial \Pi(h)}{\partial h}\bigg\rvert_{h=h_0},
\end{equation}
and
\begin{equation}\label{eq:t0}
    t_0 = \frac{3\mu}{\gamma h_0^3 q_0^4}.
\end{equation}
Solving Eq.~(\ref{eq:stability_fourier}) leads to the following expression for the spectrum~\cite{PhysRevE.100.023108,Mecke_2005},
$S(q,t) = \langle |\tilde{\delta h}(q,t)|^2\rangle$ (where the brackets stand for the average over the noise):
\begin{equation}\label{eq:structure_factor}
    S(q,t) = S_0(q)e^{2\omega(q)t} + \frac{k_BTh_0^3}{3\mu}\frac{q^2}{\omega(q)}(e^{2\omega(q)t} - 1).
\end{equation}
For $q>\sqrt{2}q_0$, $S(q,t)$ tends to the capillary wave spectrum, 
$S \sim S_{\text{cw}} \propto \frac{k_BT}{\gamma q^2}$, as $t \rightarrow \infty$ (whereas it would decay exponentially to zero in the deterministic case) ~\cite{PhysRevLett.99.114503,Mecke_2005}.
Moreover, while the maximum $q_0$ is independent of time in the deterministic case, the maximum $q_m$ of 
(\ref{eq:structure_factor}) approaches $q_0$ from the right as the time increases.
It follows that the most unstable wavelength, $\lambda_{\text{max}} = 2\pi/q_m$, grows with time.
We will come back to this fact in Sec.~\ref{sec:results_two} and show that indeed this behavior is reproduced and observed in our simulations.

\section{Thin Film Lattice Boltzmann Model}\label{sec:num_method}
Performing numerical simulations of the thin-film equation is challenging, even in the deterministic case;
including stochastic terms adds a further level of complexity.
To this aim, several sophisticated numerical methods have been developed, based on, e.g.,   
finite differences~\cite{PhysRevE.63.011208}, finite elements~\cite{Grun2006} and spectral schemes~\cite{Duran_Olivencia2019} .

@@

a class of models originally devised for the shallow water equations~\cite{Salmon:1999:0022-2402:503,PhysRevE.65.036309,zhou2004lattice,van2010study}.

@@

The equilibrium distribution functions $f_i^{(eq)}$ read~\cite{van2010study}:

@@

\subsection{Fluid-substrate interactions: the contact angle}\label{subsec:fluid_substrate}
For the disjoining pressure $\Pi(h)$ we use the following 
expression\textcolor{black}{\cite{RevModPhys.81.1131, Peschka9275, PhysRevE.92.061002, oronLongscaleEvolutionThin1997}}:
\begin{equation}\label{eq:disjoin_p}
    \Pi(h) = \kappa(\theta) \left[\left(\frac{h_{\ast}}{h+h_c}\right)^n - \left(\frac{h_{\ast}}{h+h_c}\right)^m\right],
\end{equation}
where the prefactor $\kappa(\theta) = \gamma(1-\cos(\theta))\frac{(n-1)(m-1)}{(n-m)h_{\ast}}$ is related to the Hamaker constant by
$\mathcal{A} = 6 \pi h_{\ast}^3 \kappa(\theta)$.
$\theta$ is the contact angle, which effectively encodes the interfacial energies of the three-phase system and represents, therefore, a measure of the hydrophilicity/hydrophobicity of the substrate.
\textcolor{black}{Throughout the paper we restrict ourselves to low contact angles, $\theta < 1$,
so to comply with the lubrication approximation, which requires the derivative of the height field to be small, $\partial_x h(x) \ll 1$, and hence limits
the values of $\theta$ (since, at the contact point, $\partial_x h \sim \tan(\theta) \sim \theta$).}
\textcolor{black}{To ensure numerical stability we introduce a regularizing term $h_c$; the value $h_{\ast} - h_c > 0$, 
at which the disjoining pressure vanishes, sets the thickness of the precursor film covering the substrate.}

\section{Results}\label{sec:results_two}
\begin{center}
\begin{table}
    \centering
    \begin{tabular}{||l | c | c | p{1cm} | c | c | p{1cm} ||}
        \hline
        & \multicolumn{3}{ | c | }{$n=3, m=2$} & \multicolumn{3}{ | c | }{$n=9, m=3$} \\
        \hline
         $\theta$ & $t_0$ & $q_0$ & $\lambda_{\text{max}}$ & $t_0$ & $q_0$ & $\lambda_{\text{max}}$ \\ [0.5ex] 
         \hline\hline
         $\pi/9$ & $5 \cdot 10^5$ & $0.101$ & $62$ & $8.6 \cdot 10^6$ & $0.049$ & $128$ \\ 
         \hline
         $5\pi/36$ & $2 \cdot 10^5$ & $0.126$ & $50$ & $3.5 \cdot 10^6$ & $0.061$ & $103$  \\
         \hline
         $\pi/6$ & $10^5$ & $0.151$ & $42$ & $1.7 \cdot 10^6$ & $0.073$ & $86$ \\
         \hline
         $7\pi/36$ & $5 \cdot 10^4$ & $0.175$ & $36$ & $10^6$ & $0.085$ & $74$ \\
         \hline
         $2\pi/9$ & $3 \cdot 10^4$ & $0.2$ & $32$ & $6 \cdot 10^5$ & $0.097$ & $65$ \\ [1ex] 
         \hline
        \end{tabular}
        \caption{Numerical values (in lbu) of characteristic time, $t_0$, most unstable wavenumber, $q_0$, and corresponding wavelength, $\lambda_{\text{max}}$, 
        for the various contact angles $\theta$ and pairs $(n,m)$ of the disjoining pressure exponents used (see Eq.~(\ref{eq:disjoin_p}).}
    \label{tab:t_0s}
\end{table}
\end{center}
\textcolor{black}{We simulate numerically the model 
(\ref{eq:LBE_two}-\ref{eq:hydrofields}), with forcing term (\ref{eq:capillary_force}-\ref{eq:fric_force}-\ref{eq:thermal_force}-\ref{eq:tot_force}) (and $\Delta x = \Delta t = 1$), on a domain 
of length $L=4096$,
with a relaxation time $\tau = 1$, corresponding to a kinematic viscosity $\nu = 1/6$.
The surface tension $\gamma$ is set to $0.01$ and we use, as exponents appearing in the disjoining pressure (Eq. (\ref{eq:disjoin_p})),
the pairs $(n,m)=(3,2)$ (in section \ref{subsec:validation}, \ref{subsec:morphandrup} and \ref{subsec:results_contact_angle}) 
and $(n,m)=(9,3)$ (in section \ref{subsec:patterned_substrates}).
The thickness of the precursor film is $h_{\ast}-h_c = 0.05$ ($h_{\ast}=0.1, h_c = 0.05$) and the numerical slip length $\delta$ (unless otherwise stated)
is chosen to be in the weak slip regime, $\delta/h_0 = 1.0$. % The precursor film is only 0.05 in height, as there is as well hc added to h to make up 0.1
The system is initialised with zero velocity everywhere ($u(x,0)=0 \; \forall x \in [0, L]$)
and with random perturbations of amplitude $10^{-3}$ around the value $h_0 = 1$ for the film height, i.e. 
$h(x,0) = h_0 + 10^{-3} \varepsilon$ ($\varepsilon$ being a random variable uniformly distributed 
in $[-1, +1]$).
In the remainder of the manuscript all lengths will be given adimensionalised by $h_0$.
We vary the contact angle in the interval $[\pi/9, 2\pi/9 ]$;
the numerical values (in lattice Boltzmann units, lbu) of the corresponding wavelength, $\lambda_{\text{max}}$ and wavenumber, $q_0$, of the most unstable mode as well as the characteristic time, $t_0$, can be found in Tab.~\ref{tab:t_0s}.
For each contact angle configuration we perform one simulation for the athermal system and a set 
of $O(30)$ runs for the fluctuating case, at changing the random seed (the data shown represent, then, an average over the noise realisations). We would like to stress, incidentally, that the inclusion of 
the stochastic term entails a computational overhead of, roughly, $20\%$ of a typical deterministic run.\\ 
The numerical data will be indicated, hereafter, in terms of the dimensionless temperature $\sigma = \sqrt{\frac{k_B T}{\gamma h_0^2}}$, which is set either to $\sigma=0$ (for the athermal dewetting) or to $\sigma = 3.16 \cdot 10^{-3} \equiv \sigma_0$, 
corresponding to a thermal energy
of $k_B T = 10^{-7}$ lbu (for the fluctuating dewetting). For the sake of comparison, we remark 
that such a value lies well within the window $\sigma \in [10^{-3}, 10^{-2}]$, measured in experiments with polymeric and metallic thin films~\cite{PhysRevLett.99.114503,doi:10.1021/la4009784}}.

\subsection{Testing the stochastic term}\label{subsec:validation}
\begin{figure}
    \centering
    \includegraphics[width=0.75\textwidth]{graphics/spectrum_theta_20_nob_fill_pattern.pdf}
    \caption{Height fluctuations spectra from deterministic (circles) and stochastic (triangles) simulations at $t=0.2 t_0$ (filled symbols) and $t=0.7 t_0$ (empty symbols), 
    on a substrate with $\theta =\pi/9$. The theoretical predictions, Eq.~(\ref{eq:structure_factor_2}), are reported with solid ($\sigma=0$) and dashed ($\sigma = \sigma_0$) lines.
    }
    \label{fig:theory_simulation_structure_factor}
\end{figure}
\textcolor{black}{In order to validate our model against the analytical results discussed
in section \ref{sec:stoch_thin_film}, we first rewrite Eq.~(\ref{eq:structure_factor_2}) in a form, involving the dimensionless temperature $\sigma$, which is more 
convenient for the forthcoming discussion, namely:
\begin{equation}\label{eq:structure_factor_2}
    S(q,t) = S_0 e^{2\omega(q)t} + \frac{\sigma^2 h_0^2 L}{q_0^2}g\left(\frac{q}{q_0}\right)(e^{2\omega(q)t} - 1),
\end{equation}
where $g(\xi) = \frac{\xi^2}{2\xi^2 - \xi^4}$ and the appearance of the system size $L$ stems from the noise amplitude, $\langle |\tilde{\mathcal{N}}(q,t)|^2 \rangle = L$, 
resulting from discrete Fourier-transforming 
over a finite length~\cite{PhysRevE.100.023108,PhysRevE.102.053105}. To arrive at (\ref{eq:structure_factor_2}), use was made of the expression (\ref{eq:t0}) for the characteristic time $t_0$
and we omitted the dependence on $q$ of $S_0(q)$ because, with our initialization, the latter is just a constant, $S_0(q)\equiv S_0$.}
We measured, then, the spectra $S(q,t)$, at two instants of time, 
in the early stages of the growth of the instability, for both athermal and fluctuating dewetting on a susbstrate with $\theta=\pi/9$ (and strict no-slip, $\delta=0$).
The data, reported in Fig.~\ref{fig:theory_simulation_structure_factor}, from both deterministic (circles)
and fluctuating (triangles) simulations show good agreement with the theoretical curves, Eq.~(\ref{eq:structure_factor_2}), depicted with solid ($\sigma=0$) and dashed ($\sigma=\sigma_0$) lines.
In particular, notice that the maximum  of $S(q,t)$ is attained for $q=q_0$ independently of time, in the athermal
case, and at $q = q_m > q_0$, with $q_m$ tending to $q_0$ as the time goes by, when 
fluctuations are present; in the latter case, in addition, the capillary wave spectrum $S_{\text{cw}}(q) \propto q^{-2}$
is recovered for large $q$.

>>>>

Coating processes heavily rely on the affinity between a fluid and a substrate~\cite{bonnWettingSpreading2009}.

@@

leading to film rupture~\cite{oronLongscaleEvolutionThin1997, crasterDynamicsStabilityThin2009} or wrinkling~\cite{dasilvasobrinhoStudyDefectsUltrathin1999}. 

@@

The advantage of such approach in comparison with molecule-resolved methods such as Molecular Dynamics~\cite{haile1992molecular, zhangMolecularSimulationThin2019, doi:10.1063/1.1290698, grabow1988thin}

@@

For instance, in the context of dewetting, rupture times measured in experiments are shorter than predicted by simulations of the thin-film equation~\cite{herminghaus1996dewetting,herminghaus1998dewetting,beckerComplexDewettingScenarios2003}.
More in general, thermal fluctuations can accelerate the appearance of film instabilities~\cite{Rauscher2008,tsekov1993effect, fetzerThermalNoiseInfluences2007, zhangMolecularSimulationThin2019}.
The first theoretical and experimental studies, by means of light scattering techniques, on film rupture influenced by thermal fluctuations date back to the 60s~\cite{vrijRuptureThinLiquid1968}.
Almost half a century later, a stochastic version of Eq.~(\ref{eq:thin_film}) was derived~\cite{grunThinFilmFlowInfluenced2006, meckeThermalFluctuationsThin2005, davidovitchSpreadingViscousFluid2005}, through a lubrication approximation of the Navier-Stokes-Landau-Lifshitz 
equations of fluctuating hydrodynamics~\cite{landauFluidMechanicsLandau2013}.
A strong agreement between experiments and the stochastic thin-film equation was found in showing that the variance of the interfacial roughness cannot be explained without adding a fluctuating term~\cite{fetzerThermalNoiseInfluences2007}.
The influence of thermal fluctuations on the short and long time morphology of a film has been studied with simulations of the one and two dimensional stochastic thin-film equations~\cite{nesicFullyNonlinearDynamics2015, alizadeh_pahlavan_cueto-felgueroso_hosoi_mckinley_juanes_2018}. 
Recently, new theoretical insights have been gained explaining film rupture as a consequence of the combined action of thermal fluctuations and drainage, both numerically~\cite{shahThermalFluctuationsCapillary2019} and experimentally~\cite{PhysRevLett.125.158001}.

@@

\section{Linear Stability Analysis of the Stochastic Thin-Film Equation}\label{sec:stoch_thin_film}
The one-dimensional stochastic thin-film equation reads~\cite{grunThinFilmFlowInfluenced2006, meckeThermalFluctuationsThin2005}:
\begin{equation}\label{eq:stoch_thin_film_simple}
    \partial_t h = \partial_x \left[\frac{h^3}{3\mu}\partial_x p + \sqrt{\frac{2k_BTh^3}{3\mu}}\mathcal{N}\right],
\end{equation}
where $k_B$ is Boltzmann's constant, $T$ is the temperature and $\mathcal{N}$ is a Gaussian white noise with
\begin{align}\label{eq:correlation}
    \langle\mathcal{N}(x,t)\rangle =&~0,\\ 
    \quad \langle\mathcal{N}(x,t)\mathcal{N}(x',t')\rangle =&~\delta(x-x')\delta(t-t').
\end{align}
In order to understand the influence of thermal fluctuations on the temporal evolution of the film height, we cursory recall the linear stability analysis of Eq.~(\ref{eq:stoch_thin_film_simple}) 
\cite{zhangMolecularSimulationThin2019, diezMetallicthinfilmInstabilitySpatially2016}.
Introducing the deviation $\delta h$ from the mean height $h_0$, such that $h = h_0 + \delta h$ and $\delta h \ll h_0$, 
the linearised stochastic thin-film equation is obtained
\begin{equation}\label{eq:linearstability_realspace}
    \partial_t \delta h = \frac{h_0^3}{3\mu}\Big(\Pi^{\prime}(h_0)\partial_x^2\delta h - \gamma \partial_x^4\delta h\Big) + \sqrt{\frac{2k_BT h_0^3}{3\mu}} \partial_x \mathcal{N},
\end{equation}
where $\Pi^{\prime}(h_0) \equiv \frac{\partial \Pi(h)}{\partial h}\bigg\rvert_{h=h_0}$.
Here, it is assumed that also the noise is small, in the sense that $k_B T \ll \gamma h_0^2$.
To derive a dispersion relation for this system, first the Fourier transforms of the perturbation $\delta h$  
\begin{equation}\label{eq:fourier_delta_h}
    \tilde{\delta h}(q,t) = \int_{-\infty}^{\infty} \delta h(x,t) e^{-ixq} dx,
\end{equation}
and the noise term $\mathcal{N}$
\begin{equation}\label{eq:fourier_noise}
    %\tilde{\mathcal{N}}(\mathbf{q},t) = \int_{-\infty}^{\infty} \mathcal{N}(\mathbf{x},t) e^{-i\mathbf{x}\cdot\mathbf{q}} \dd[2]{\mathbf{x}}.
    \tilde{\mathcal{N}}(q,t) = \int_{-\infty}^{\infty} \mathcal{N}(x,t) e^{-ixq} dx.
\end{equation}
must be performed.

Inserting $\tilde{\delta h}$ and $\tilde{\mathcal{N}}$ in~(\ref{eq:linearstability_realspace}) yields
\begin{equation}\label{eq:stability_fourier}
    \partial_t \tilde{\delta h} = \omega(q)\tilde{\delta h} +i\sqrt{\frac{2k_BTh_0^3}{3\mu}}q\tilde{\mathcal{N}},
\end{equation}
with dispersion relation $\omega(q)$ given by
\begin{equation}\label{eq:dispersion}
    \omega(q) = \frac{1}{t_0}\left[2\left(\frac{q}{q_0}\right)^2 - \left(\frac{q}{q_0}\right)^4\right].
\end{equation}
The maximum, $q_0$, of $\omega(q)$ and the characteristic time, $t_0$, are specific to the substrate the film is deposited on and their expressions are~\cite{fetzerThermalNoiseInfluences2007}
\begin{equation}\label{eq:q0}
    q_0^2 = \frac{1}{2\gamma}\frac{\partial \Pi(h)}{\partial h}\bigg\rvert_{h=h_0},
\end{equation}
and
\begin{equation}\label{eq:t0}
    t_0 = \frac{3\mu}{\gamma h_0^3 q_0^4}.
\end{equation}
Solving Eq.~(\ref{eq:stability_fourier}) leads to the following expression for the spectrum~\cite{zhangMolecularSimulationThin2019,meckeThermalFluctuationsThin2005},
$S(q,t) = \langle |\tilde{\delta h}(q,t)|^2\rangle$ (where the brackets stand for the average over the noise):
\begin{equation}\label{eq:structure_factor}
    S(q,t) = S_0(q)e^{2\omega(q)t} + \frac{k_BTh_0^3}{3\mu}\frac{q^2}{\omega(q)}(e^{2\omega(q)t} - 1).
\end{equation}
For $q>\sqrt{2}q_0$, $S(q,t)$ tends to the capillary wave spectrum, 
$S \sim S_{\text{cw}} \propto \frac{k_BT}{\gamma q^2}$, as $t \rightarrow \infty$ (whereas it would decay exponentially to zero in the deterministic case) ~\cite{fetzerThermalNoiseInfluences2007,meckeThermalFluctuationsThin2005}.
Moreover, while the maximum $q_0$ is independent of time in the deterministic case, the maximum $q_m$ of 
(\ref{eq:structure_factor}) approaches $q_0$ from the right as the time increases.
It follows that the most unstable wavelength, $\lambda_{\text{max}} = 2\pi/q_m$, grows with time.
We will come back to this fact in Sec.~\ref{sec:results_two} and show that indeed this behavior is reproduced and observed in our simulations.

\section{Thin Film Lattice Boltzmann Model}\label{sec:num_method}
Performing numerical simulations of the thin-film equation is challenging, even in the deterministic case;
including stochastic terms adds a further level of complexity.
To this aim, several sophisticated numerical methods have been developed, based on, e.g.,   
finite differences~\cite{diezGlobalModelsMoving2000}, finite elements~\cite{grunThinFilmFlowInfluenced2006} and spectral schemes~\cite{duran-olivenciaInstabilityRuptureFluctuations2019} .

@@

a class of models originally devised for the shallow water equations~\cite{salmonLatticeBoltzmannMethod1999,dellarNonhydrodynamicModesPriori2002,zhouLatticeBoltzmannMethods2004,vanthangStudy1DLattice2010}.

@@

The equilibrium distribution functions $f_i^{(eq)}$ read~\cite{vanthangStudy1DLattice2010}:

@@

\subsection{Fluid-substrate interactions: the contact angle}\label{subsec:fluid_substrate}
For the disjoining pressure $\Pi(h)$ we use the following 
expression\textcolor{black}{\cite{crasterDynamicsStabilityThin2009, peschkaSignaturesSlipDewetting2019, nesicFullyNonlinearDynamics2015, oronLongscaleEvolutionThin1997}}:
\begin{equation}\label{eq:disjoin_p}
    \Pi(h) = \kappa(\theta) \left[\left(\frac{h_{\ast}}{h+h_c}\right)^n - \left(\frac{h_{\ast}}{h+h_c}\right)^m\right],
\end{equation}
where the prefactor $\kappa(\theta) = \gamma(1-\cos(\theta))\frac{(n-1)(m-1)}{(n-m)h_{\ast}}$ is related to the Hamaker constant by
$\mathcal{A} = 6 \pi h_{\ast}^3 \kappa(\theta)$.
$\theta$ is the contact angle, which effectively encodes the interfacial energies of the three-phase system and represents, therefore, a measure of the hydrophilicity/hydrophobicity of the substrate.
\textcolor{black}{Throughout the paper we restrict ourselves to low contact angles, $\theta < 1$,
so to comply with the lubrication approximation, which requires the derivative of the height field to be small, $\partial_x h(x) \ll 1$, and hence limits
the values of $\theta$ (since, at the contact point, $\partial_x h \sim \tan(\theta) \sim \theta$).}
\textcolor{black}{To ensure numerical stability we introduce a regularizing term $h_c$; the value $h_{\ast} - h_c > 0$, 
at which the disjoining pressure vanishes, sets the thickness of the precursor film covering the substrate.}

\section{Results}\label{sec:results_two}
\begin{center}
\begin{table}
    \centering
    \begin{tabular}{||l | c | c | p{1cm} | c | c | p{1cm} ||}
        \hline
        & \multicolumn{3}{ | c | }{$n=3, m=2$} & \multicolumn{3}{ | c | }{$n=9, m=3$} \\
        \hline
         $\theta$ & $t_0$ & $q_0$ & $\lambda_{\text{max}}$ & $t_0$ & $q_0$ & $\lambda_{\text{max}}$ \\ [0.5ex] 
         \hline\hline
         $\pi/9$ & $5 \cdot 10^5$ & $0.101$ & $62$ & $8.6 \cdot 10^6$ & $0.049$ & $128$ \\ 
         \hline
         $5\pi/36$ & $2 \cdot 10^5$ & $0.126$ & $50$ & $3.5 \cdot 10^6$ & $0.061$ & $103$  \\
         \hline
         $\pi/6$ & $10^5$ & $0.151$ & $42$ & $1.7 \cdot 10^6$ & $0.073$ & $86$ \\
         \hline
         $7\pi/36$ & $5 \cdot 10^4$ & $0.175$ & $36$ & $10^6$ & $0.085$ & $74$ \\
         \hline
         $2\pi/9$ & $3 \cdot 10^4$ & $0.2$ & $32$ & $6 \cdot 10^5$ & $0.097$ & $65$ \\ [1ex] 
         \hline
        \end{tabular}
        \caption{Numerical values (in lbu) of characteristic time, $t_0$, most unstable wavenumber, $q_0$, and corresponding wavelength, $\lambda_{\text{max}}$, 
        for the various contact angles $\theta$ and pairs $(n,m)$ of the disjoining pressure exponents used (see Eq.~(\ref{eq:disjoin_p}).}
    \label{tab:t_0s}
\end{table}
\end{center}
\textcolor{black}{We simulate numerically the model 
(\ref{eq:LBE_two}-\ref{eq:hydrofields}), with forcing term (\ref{eq:capillary_force}-\ref{eq:fric_force}-\ref{eq:thermal_force}-\ref{eq:tot_force}) (and $\Delta x = \Delta t = 1$), on a domain 
of length $L=4096$,
with a relaxation time $\tau = 1$, corresponding to a kinematic viscosity $\nu = 1/6$.
The surface tension $\gamma$ is set to $0.01$ and we use, as exponents appearing in the disjoining pressure (Eq. (\ref{eq:disjoin_p})),
the pairs $(n,m)=(3,2)$ (in section \ref{subsec:validation}, \ref{subsec:morphandrup} and \ref{subsec:results_contact_angle}) 
and $(n,m)=(9,3)$ (in section \ref{subsec:patterned_substrates}).
The thickness of the precursor film is $h_{\ast}-h_c = 0.05$ ($h_{\ast}=0.1, h_c = 0.05$) and the numerical slip length $\delta$ (unless otherwise stated)
is chosen to be in the weak slip regime, $\delta/h_0 = 1.0$. % The precursor film is only 0.05 in height, as there is as well hc added to h to make up 0.1
The system is initialised with zero velocity everywhere ($u(x,0)=0 \; \forall x \in [0, L]$)
and with random perturbations of amplitude $10^{-3}$ around the value $h_0 = 1$ for the film height, i.e. 
$h(x,0) = h_0 + 10^{-3} \varepsilon$ ($\varepsilon$ being a random variable uniformly distributed 
in $[-1, +1]$).
In the remainder of the manuscript all lengths will be given adimensionalised by $h_0$.
We vary the contact angle in the interval $[\pi/9, 2\pi/9 ]$;
the numerical values (in lattice Boltzmann units, lbu) of the corresponding wavelength, $\lambda_{\text{max}}$ and wavenumber, $q_0$, of the most unstable mode as well as the characteristic time, $t_0$, can be found in Tab.~\ref{tab:t_0s}.
For each contact angle configuration we perform one simulation for the athermal system and a set 
of $O(30)$ runs for the fluctuating case, at changing the random seed (the data shown represent, then, an average over the noise realisations). We would like to stress, incidentally, that the inclusion of 
the stochastic term entails a computational overhead of, roughly, $20\%$ of a typical deterministic run.\\ 
The numerical data will be indicated, hereafter, in terms of the dimensionless temperature $\sigma = \sqrt{\frac{k_B T}{\gamma h_0^2}}$, which is set either to $\sigma=0$ (for the athermal dewetting) or to $\sigma = 3.16 \cdot 10^{-3} \equiv \sigma_0$, 
corresponding to a thermal energy
of $k_B T = 10^{-7}$ lbu (for the fluctuating dewetting). For the sake of comparison, we remark 
that such a value lies well within the window $\sigma \in [10^{-3}, 10^{-2}]$, measured in experiments with polymeric and metallic thin films~\cite{fetzerThermalNoiseInfluences2007,doi:10.1021/la4009784}}.

\subsection{Testing the stochastic term}\label{subsec:validation}
\begin{figure}
    \centering
    \includegraphics[width=0.75\textwidth]{graphics/spectrum_theta_20_nob_fill_pattern.pdf}
    \caption{Height fluctuations spectra from deterministic (circles) and stochastic (triangles) simulations at $t=0.2 t_0$ (filled symbols) and $t=0.7 t_0$ (empty symbols), 
    on a substrate with $\theta =\pi/9$. The theoretical predictions, Eq.~(\ref{eq:structure_factor_2}), are reported with solid ($\sigma=0$) and dashed ($\sigma = \sigma_0$) lines.
    }
    \label{fig:theory_simulation_structure_factor}
\end{figure}
\textcolor{black}{In order to validate our model against the analytical results discussed
in section \ref{sec:stoch_thin_film}, we first rewrite Eq.~(\ref{eq:structure_factor_2}) in a form, involving the dimensionless temperature $\sigma$, which is more 
convenient for the forthcoming discussion, namely:
\begin{equation}\label{eq:structure_factor_2}
    S(q,t) = S_0 e^{2\omega(q)t} + \frac{\sigma^2 h_0^2 L}{q_0^2}g\left(\frac{q}{q_0}\right)(e^{2\omega(q)t} - 1),
\end{equation}
where $g(\xi) = \frac{\xi^2}{2\xi^2 - \xi^4}$ and the appearance of the system size $L$ stems from the noise amplitude, $\langle |\tilde{\mathcal{N}}(q,t)|^2 \rangle = L$, 
resulting from discrete Fourier-transforming 
over a finite length~\cite{zhangMolecularSimulationThin2019,zhangNanoscaleThinfilmFlows2020}. To arrive at (\ref{eq:structure_factor_2}), use was made of the expression (\ref{eq:t0}) for the characteristic time $t_0$
and we omitted the dependence on $q$ of $S_0(q)$ because, with our initialization, the latter is just a constant, $S_0(q)\equiv S_0$.}
We measured, then, the spectra $S(q,t)$, at two instants of time, 
in the early stages of the growth of the instability, for both athermal and fluctuating dewetting on a susbstrate with $\theta=\pi/9$ (and strict no-slip, $\delta=0$).
The data, reported in Fig.~\ref{fig:theory_simulation_structure_factor}, from both deterministic (circles)
and fluctuating (triangles) simulations show good agreement with the theoretical curves, Eq.~(\ref{eq:structure_factor_2}), depicted with solid ($\sigma=0$) and dashed ($\sigma=\sigma_0$) lines.
In particular, notice that the maximum  of $S(q,t)$ is attained for $q=q_0$ independently of time, in the athermal
case, and at $q = q_m > q_0$, with $q_m$ tending to $q_0$ as the time goes by, when 
fluctuations are present; in the latter case, in addition, the capillary wave spectrum $S_{\text{cw}}(q) \propto q^{-2}$
is recovered for large $q$.
